A user-friendly tool to evaluate the effectiveness of no-take marine reserves

### 2017-12-15 Manuscript.Rmd
updated manuscript with comments from Melaina

---
title: A user--friendly tool to evaluate the effectiveness of no--take marine reserves
subtitle: Marine reserve evaluation
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
fontsize: 12pt
bibliography:
- exported-references.bib
- packages.bib
csl: plos.csl
---
<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { 
      equationNumbers: { 
            autoNumber: "all",
            formatNumber: function (n) {return '9.'+n}
      } 
  }
});
</script>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
options(knitr.table.format = "latex")

@@

Marine reserves are implemented to achieve a variety of objectives, but are seldom rigorously evaluated to determine whether those objectives are met. In the rare cases when evaluations do take place, they typically focus on ecological indicators and ignore other relevant objectives such as socioeconomics and governance. And regardless of the objectives, the diversity of locations, monitoring protocols, and analysis approaches hinder the ability to compare results across case studies. Moreover, analysis and evaluation of reserves is generally conducted by outside researchers, not the reserve managers or users, plausibly thereby hindering effective local management and rapid response to change. We present a framework and tool, called “MAREA”, to overcome these challenges. Its purpose is to evaluate the extent to which any given reserve has achieved its stated objectives.  MAREA provides specific guidance on data collection and formatting, and then conducts rigorous causal inference analysis based on data input by the user, providing real--time outputs about the effectiveness of the reserve. MAREA’s ease of use, standardization of state--of--the--art inference methods, and ability to analyze marine reserve effectiveness across ecological, socioeconomic, and governance objectives could dramatically further our understanding and support of effective marine reserve management.

@@

Empirical evidence shows that MPAs increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience to climatic impacts [@micheli_2012-EU; @roberts_2017-J9], and preserve genetic diversity [@munguavega_2015-yg]. Compared to MPAs that grant partial protection, marine reserves have higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_; @edgar_2014-UO; @giakoumi_2017-V2; @sala_2017-69]. However, these effects are often measured as biological changes within the reserves through time, and many lack a control site for comparison [@betti_2017-lq]. This approach does not account for other factors (*e.g.* system-level changes in productivity caused by predatory release @szuwalski_2017-jc; or favorable environmental conditions @chavez_2003-_m) for which one must control [@davies_2017-ml] in order to causally attribute a biological change to the reserve. Other studies have used a control--impact comparison approach that uses control sites but does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].

A smaller fraction of studies have used a before--after--control--impact (*i.e.* BACI) design comparing reserves to control sites before and after implementation [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], which allows the use of causal inference techniques that estimate the effect of the reserve. For example, in ref [@moland_2013-VP] authors use a BACI design and observe increases in lobster catches --a proxy for abundances-- after the implementation of the reserves for protected and control sites. However, the temporal changes in the reserve were greater than in the control site, suggesting a positive effect off the reserve on lobster catches. But even when proper causal inference can be drawn, results are often different across reserves. Effects of reserves on ecological and economic outcomes are highly heterogeneous, and often depend on the specific ecological, economic, and social context.

@@

An increasingly popular way to make science accessible, reproducible, scalable, and replicable is through Open Science and the development of open--access tools [@lowndes_2017-xh]. The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, successfully standardized a way to measure the health and benefits of the oceans. This approach has been implemented at a global scale, but also at country-level [@selig_2015-F9], and regionally [@halpern_2014-lQ; @elfes_2014-RC]. Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance [@anderson_2015-ND; @dowling_2016-pO], design territorial use rights for fisheries [TURFs; @oyanedel_2017-TO], and improve decision making in the hydro power industry [@vilela_2017-Zo].

The purpose of this paper is to describe a user--friendly tool, called “MAREA”, to rigorously systematize the evaluation of marine reserve effectiveness in terms of fisheries and marine conservation. The tool is in the form of an open--source application that uses state--of--the--art methods from program evaluation to compare a reserve to control sites along a number of biological, economic, and governance dimensions. We first provide a list of commonly stated management objectives and match them to appropriate indicators. We then develop a simple approach to analyzing these indicators building on causal inference techniques [@moland_2013-VP], which help us understand the effect of management interventions [@burgess_2016-HN; @mascia_2017-m_]. To implement the analytical approach, we introduce the Marine Reserve Evaluation Application (MAREA), an open source, web–based tool that automates the framework described in this paper and enables its broader use. Finally, we present a case study on the evaluation of a marine reserve established by the fishers of Isla Natividad (Mexico) in 2006, to demonstrate the potential of MAREA.

@@

Here, we describe the proposed framework to evaluate the effectiveness of marine reserves (Fig. 1). We explain how management objectives were identified and matched to appropriate indicators that allow the evaluation of the reserves, and provide brief guidelines on data collection. Alongside, methodologies to analyze these indicators are presented. We then describe the development of MAREA and explain how this open--access tool can be used by fishermen, managers, and other stakeholders with little scientific background. Finally, we provide guidelines on how to interpret and use the results and output generated by MAREA to inform management.

@@

Throughout this study, we will refer to the stated goals for which a marine reserve was designed as “objectives.” This work was motivated by the case of Mexico, where 39 reserves have been implemented over the past five years to achieve objectives such as increasing productivity in nearby waters or recovery of overexploited species; most of these reserves have never been formally evaluated for effectiveness at meeting those objectives. Thus, our focus was on identifying common objectives of marine reserves in Mexico. However, a literature review and discussions with marine reserve researchers suggest that the objectives driving Mexican marine reserve implementation are similar to those in the rest of the world. Thus, we group these objectives into seven major categories that may be applied to marine reserves worldwide. Any given reserve may have been implemented to meet one or more of these. The list includes objectives stated in legislation [@nom049sagpesc_2014-V6; @lgeepa_2017-jL] and official documents such as the Technical Justification Studies (*Estudios Técnicos Justificativos*), agreements, and decrees associated to these areas:

@@

Based on these seven objectives, we determined a set of associated indicators to evaluate reserve effectiveness. These indicators are specific variables on which data could be collected, and analyzed, to ultimately determine whether the corresponding objective was causally being achieved by the marine reserve. The list of indicators was compiled through a review of scientific literature in which we identified indicators that were used to measure similar objectives[@lester_2017-nh; @sala_2017-69; @chirico_2017-Rz; @edgar_2014-UO; @rodriguez_2017-PD; @rossetto_2015-V0; @betti_2017-lq; @sala_2016-PV; @woodcock_2017-Wm; @friedlander_2017-oI; @moland_2013-VP; @aburtooropeza_2011-ya; @guidetti_2014-8Z; @lester_2008-F_; @lester_2009-Ks; @pomeroy_2005-Py; @pomeroy_2004-23]. A first filter eliminated indicators for which baseline data do not typically exist in Mexico. The preliminary list of indicators was reviewed at a workshop with participation of members from Mexican fishery management agencies and non-government organizations. Later, these were presented to fishers from the Ensenada Fishing Cooperative (*S.C.P.P. Ensenada*), in El Rosario, Baja California, who provided input. Our final list of indicators includes those identified in review works [@lester_2009-Ks; @woodcock_2017-Wm].

Indicators are divided into three main categories: biological, socioeconomic, and governance (Table 1). The nine biological indicators focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data and Analysis section for specific sampling design and methodologies). Five socioeconomic indicators reflect the performance of the fishery in terms of landings, income from landings, and availability of alternative livelihoods. Fifteen governance indicators describe the governance structures under which the community operates (*e.g.*, access rights to the fishery, number of fishers, legal recognition of the reserve). Most biological and socioeconomic indicators are quantitative and require a numerical entry (*e.g.* Fish biomass) while all governance indicators, one biological indicator, and one socioeconomic indicator are qualitative and rely on a descriptive entry (*e.g.* Reasoning for reserve location). Many of them specifically measure an outcome of the reserve, though some are designed to further the understanding of the mechanisms driving a reserve’s performance. In that sense, most biological and socioeconomic indicators are outcome variables. On the other hand, governance indicators are viewed as possible explanatory variables of reserve performance. Whenever an indicator is applied to “Target species”, it means that the indicator can be used for all species (*e.g.* Fish Biomass) and/or for individual species that are either the conservation target  of the reserve or are of particular economic or ecological interest (*e.g.* Grouper Biomass). Finally, indicators B3 and B4 are different in that B3 only looks at the density of organisms above size at first maturity (related to reproductive potential), while B4 measures the density of all fish or of a target species. Each indicator targets different plausible desired outcomes, like increased reproductive potential (*i.e.* B3; @carter_2017-Uf) or having more fish -regardless of their size- to attract tourism (*i.e.* B4). Table 1 presents the proposed indicators, and Table 2 shows how objectives are matched with biological and socioeconomic indicators. Governance indicators are excluded from Table 2, as governance indicators are important to consider for every objective.

```{r, results = "asis"}
read.csv("./Tables/Table_indicators.csv") %>% 
  knitr::kable(caption = "List of indicators to evaluate the effectiveness of no-take marine reserves.",
               col.names = c("Code", "Indicator", "Data type", "Unit")) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  group_rows("Biological", 1, 7) %>%
  group_rows("Socioeconomic", 8, 10) %>% 
  group_rows("Governance", 11, 25) %>% 
  add_footnote(c("The indicator is also applied to target species"), notation = "symbol")
```

@@

```{r, results = "asis"}
read.csv("./Tables/Objectives_Indicators.csv") %>% 
  select(1:15) %>% 
  knitr::kable(caption = "Management objectives and respective performance indicators.",
               col.names = c("Objective", "B1", "B2", "B3", "B4", "B4*", "B5", "B6", "B7", "B7*", "S1", "S1*", "S2", "S2*", "S3")) %>% 
  kable_styling(latex_options = c("scale_down"))  %>% 
  column_spec(1, width = "4cm") %>% 
  add_footnote(c("The indicator is applied to target species"), notation = "symbol")
```

@@

Socioeconomic data are often collected by fishers, natural resource management agencies, or CSOs by recording landings, income, and sometimes prices for each species. To control for inflation, income is adjusted with the country’s consumer price index [@oecd_website_2017-VV]:

@@

While this model does not formally allow for causal inference, we can still measure changes in mean landings and income before and after the implementation of the reserve and provide valuable input. For both models (eq. 1 and eq. 3), we estimate the model coefficient with ordinary least squares, and calculate heteroskedastic–robust standard errors [@sandwich_2014].

@@

We developed MAREAin R version 3.4.2 and R Studio 1.1.383 [@rcore_2017] using the Shiny package [@shiny_2017], to build an interactive web application hosted on an open server; the MAREA app can be accessed at [turfeffect.shinyapps.io/marea](turfeffect.shinyapps.io/marea/). While the original version was developed in Spanish because it was aimed for Mexico and other Latin-American countries, all of its content can be translated by a translation widget available within the app.

MAREA is designed as a 6--step process, divided in tabs that appear upon launching the app. The first tab introduces the app and summarizes the evaluation process. Then, the user selects management objectives, which MAREA automatically matches to appropriate indicators, based on Table 2. Users can also select and deselect indicators based on their interests and data availability by "clicking" on the check-boxes in MAREA. The user can then load data on one or more reserves, using standard \*.csv text files; sample datasets are provided within MAREA. Once data have been loaded, MAREA identifies all reserves in the data, and lets the user select the reserve to be evaluated. At this point, the user can also specify the year of implementation of the reserve, reserve dimensions, and indicate target species that are of particular management interest. MAREA provides the user with a section to confirm that all the decisions made leading up to that point are correct. Once the user has confirmed all input data, objectives, and other information, MAREA performs the formal program evaluation analyses discussed above. For a typical data set, the automated analysis step takes less than one second. Finally, the user is taken to the results tab where all results are presented in a simple format. The user can also download a more comprehensive technical report produced in \*.pdf  format.

@@

**Fig 3. General location of Isla Natividad (left) and map of the island (right).** The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce). Shapefiles for Mexican coastline and the US were obtained from the Instituto Nacional de Estadistica, Geografia e Informatica of Mexico (INEGI: www.inegi.org.mx/geo/contenidos/geoestadistica/m_geoestadistico.aspx) and the tmap R package [@tmap_2017], respectively.

@@

In 2006, the Isla Natividad community established a biological baseline following the data collection protocol described in this study. The community then implemented two community--based marine reserves within their TURF [@afflerbach_2014-HP; @lester_2017-nh; @micheli_2012-EU] after establishing a baseline for the soon--to--be reserves and control sites. Evidence suggest that these reserves have been effective at enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of high value commercial species such as abalone [@munguavega_2015-yg]. These ecological benefits have been translated into economic benefits, enhancing population persistence and bolstering abalone fisheries [@rossetto_2015-V0]. For the purpose of this evaluation, we focused on the “La Plana / Las Cuevas” marine reserve, located at the southern end of the island (Fig. 3) and its corresponding control site “La Dulce / Babencho”.

@@

Socioeconomic data were obtained from the National Commission for Aquaculture and Fisheries (*Comisión Nacional de Acuacultura y Pesca*; CONAPESCA). The data contains species--level information on monthly landings and income from nine species from 2000 to 2014. Data on landings and income were aggregated by year and species, and adjusted by the Consumer Price Index [@oecd_website_2017-VV]. From the nine species available, we selected as objective species those that contributed the most (88.27%) income from 2000 to 2014: lobster (*Panulirus interruptus*; 71.76%), red sea urchin (*Mesocentrotus franciscanus*; 9.33%), snail (*Megastraea undosa*; 3.93%), and sea cucumber (*Parastichopus parvimensis*; 3.23%). Abalone species (*Haliotis fulgens*; 4.52% and *Haliotis corrugata*; 6.16%) were excluded because the cooperative implemented an informal closure of these fisheries in 2010 to allow the population recover. Eliminating all fishing pressure on abalones means that the control site receives (for this species) the same treatment as the reserve.

@@

```{r table of bio results, results = "asis"}
bio_results_summary %>% 
  mutate(estimate = paste0(formatC(estimate, digits = 2, format = "f"),
                           " (",
                           formatC(std.error, digits = 2, format = "f"),
                           ")",
                           significance)) %>% 
  select(Indicator, estimate, tscore = statistic) %>% 
  kable(caption = "Summary of average treatment effect of the reserve on biological indicators.",
        digits = c(NA, NA, 4),
        col.names = c("Indicator", "Estimate (SD)", "t-score")) %>% 
  kable_styling() %>% 
  add_footnote(c("Asterisks indicate significance level, with (*) indicating p < 0.1 and (**) p < 0.05."),
               notation = "symbol")
```

One of the main objectives of this reserve was to increase landings. Results of the socioeconomic indicators show that total landings were, on average, 64.20 metric tonnes higher (*p* > 0.1) after the implementation of the reserves, though this cannot necessarily be interpreted as causal, because it relies entirely on a before--after comparison. Total income was $10,344.85 (*p* < 0.05) thousands of Mexican Pesos (K MXP) higher after the implementation of the reserves. On average, lobster and sea cucumber landings increased, while urchin and snail landings and income decreased. Figure 6 presents the changes in these indicators through time, and Table 4 summarizes these results.

@@

```{r table of soc results, results = "asis"}
socio_results_summary %>% 
  mutate(estimate = paste0(formatC(estimate, digits = 2, format = "f"),
                           " (",
                           formatC(std.error, digits = 2, format = "f"),
                           ")",
                           significance)) %>% 
  select(Indicator, estimate, tscore = statistic) %>% 
  kable(caption = "Summary of differences in socioeconomic indicators before and after the implementation of the reserve.",
        digits = c(NA, NA, 4),
        col.names = c("Indicator", "Estimate (SD)", "t-score")) %>% 
  kable_styling() %>% 
  add_footnote(c("Asterisks indicate significance level, with (*) indicating p < 0.1 and (**) p < 0.05."),
               notation = "symbol")
```

Recall that the governance objectives are evaluated based on the institutions present, not on a specific quantitative linkage between governance and biological or economic outcomes. Data for this reserve suggest that the community is strongly organized, which is a likely driver of the successes reported above [@gutirrez_2011-0U]. The first point of success is the existence of a fishing cooperative that is also affiliated with a regional federation of cooperatives. These polycentric governance structures allow various levels of organization that have been shown to foster communication and cooperation [@mccay_2014-CN; @espinosaromero_2014-PY]; federations also provide bargain power with governments [@espinosaromero_2014-PY; @finkbeiner_2015-87]. Access to fishing resources is managed through a TURF, permits, and fishing quotas (for some species). McCay [@mccay_2017-1m] suggests that the TURF promotes a sense of stewardship of their resources and incentivizes sustainable management. Together, these structures enabled a participative, bottom–up process during the reserve design phase; opinions of all fishing members ---and often non--fishing community members--- were included. Participation of community members in reserve surveillance and yearly monitoring indicate commitment and interest, and allow informal communication of results to un--involved community members. Furthermore, the reserve is partially isolated from poaching activity and fishers have internal regulations pertaining to the reserves. The low level of illegal fishing by members of the community and outsiders both inside and outside the reserve represents another indication of effectiveness. Governance indicators are summarized in Table 5.

>>>>

---
title: A user-friendly tool to evaluate the effectiveness of no-take marine reserves
subtitle: Marine reserve evaluation
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
fontsize: 12pt
bibliography:
- exported-references.bib
- packages.bib
csl: plos.csl
---
<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { 
      equationNumbers: { 
            autoNumber: "all",
            formatNumber: function (n) {return '9.'+n}
      } 
  }
});
</script>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
options(knitr.table.format = "latex")

@@

Marine reserves are implemented to achieve a variety of objectives, but are seldom rigorously evaluated to determine whether those objectives are met. In the rare cases when evaluations do take place, they typically focus on ecological indicators and ignore other relevant objectives such as socioeconomics and governance. And regardless of the objectives, the diversity of locations, monitoring protocols, and analysis approaches hinder the ability to compare results across case studies. Moreover, analysis and evaluation of reserves is generally conducted by outside researchers, not the reserve managers or users, plausibly thereby hindering effective local management and rapid response to change. We present a framework and tool, called “MAREA”, to overcome these challenges. Its purpose is to evaluate the extent to which any given reserve has achieved its stated objectives.  MAREA provides specific guidance on data collection and formatting, and then conducts rigorous causal inference analysis based on data input by the user, providing real-time outputs about the effectiveness of the reserve. MAREA’s ease of use, standardization of state-of-the-art inference methods, and ability to analyze marine reserve effectiveness across ecological, socioeconomic, and governance objectives could dramatically further our understanding and support of effective marine reserve management.

@@

Empirical evidence shows that MPAs increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience to climatic impacts [@micheli_2012-EU; @roberts_2017-J9], and preserve genetic diversity [@munguavega_2015-yg]. Compared to MPAs that grant partial protection, marine reserves have higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_; @edgar_2014-UO; @giakoumi_2017-V2; @sala_2017-69]. However, these effects are often measured as biological changes within the reserves through time, and many lack a control site for comparison [@betti_2017-lq]. This approach does not account for other factors (*e.g.* system-level changes in productivity caused by predatory release @szuwalski_2017-jc; or favorable environmental conditions @chavez_2003-_m) for which one must control [@davies_2017-ml] in order to causally attribute a biological change to the reserve. Other studies have used a control-impact comparison approach that uses control sites but does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].

A smaller fraction of studies have used a before-after-control-impact (*i.e.* BACI) design comparing reserves to control sites before and after implementation [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], which allows the use of causal inference techniques that estimate the effect of the reserve. For example, in ref [@moland_2013-VP] authors use a BACI design and observe increases in lobster catches --a proxy for abundances-- after reserve implementation for protected and control sites. However, the temporal changes in the reserve were greater than in the control site, suggesting a positive effect of the reserve on lobster catches. But even when proper causal inference can be drawn, results are often different across reserves. Effects of reserves on ecological and economic outcomes are highly heterogeneous, and often depend on the specific ecological, economic, and social context.

@@

An increasingly popular way to make science accessible, reproducible, scalable, and replicable is through Open Science and the development of open-access tools [@lowndes_2017-xh]. The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, successfully standardized a way to measure the health and benefits of the oceans. This approach has been implemented at a global scale, but also at country-level [@selig_2015-F9], and regionally [@halpern_2014-lQ; @elfes_2014-RC]. Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance [@anderson_2015-ND; @dowling_2016-pO], design territorial use rights for fisheries [TURFs; @oyanedel_2017-TO], and improve decision making in the hydro power industry [@vilela_2017-Zo].

The purpose of this paper is to describe a user-friendly tool, called “MAREA”, to rigorously systematize the evaluation of marine reserve effectiveness in terms of fisheries and marine conservation goals. The tool is in the form of an open-source application that uses state-of-the-art methods from program evaluation to compare a reserve to control sites along a number of biological, economic, and governance dimensions. We first provide a list of commonly stated management objectives and match them to appropriate indicators. We then develop a simple approach to analyzing these indicators building on causal inference techniques [@moland_2013-VP], which help us understand the effect of management interventions [@burgess_2016-HN; @mascia_2017-m_]. To implement the analytical approach, we introduce the Marine Reserve Evaluation Application (MAREA), an open source, web–based tool that automates the framework described in this paper and enables its broader use. Finally, we present a case study on the evaluation of a marine reserve established by the fishers of Isla Natividad (Mexico) in 2006, to demonstrate the potential of MAREA.

@@

Here, we describe the proposed framework to evaluate the effectiveness of marine reserves (Fig. 1). We explain how management objectives were identified and matched to appropriate indicators that allow the evaluation of the reserves, and provide brief guidelines on data collection. Alongside, methodologies to analyze these indicators are presented. We then describe the development of MAREA and explain how this tool can be used by fishermen, managers, and other stakeholders with little scientific background. Finally, we provide guidelines on how to interpret and use the results and output generated by MAREA to inform management.

@@

Throughout this study, we will refer to the stated goals for which a marine reserve was designed as “objectives.” This work was motivated by the case of Mexico, where 39 reserves have been implemented over the past five years to achieve objectives such as increasing productivity in nearby waters or recovery of overexploited species; most of these reserves have never been formally evaluated for effectiveness at meeting those objectives. Thus, our focus was on identifying common objectives of marine reserves in Mexico. However, a literature review and discussions with marine reserve researchers suggested that the objectives driving Mexican marine reserve implementation are similar to those in the rest of the world. Thus, we group these objectives into seven major categories that may be applied to marine reserves worldwide. Any given reserve may have been implemented to meet one or more of these. The list includes objectives stated in legislation [@nom049sagpesc_2014-V6; @lgeepa_2017-jL] and official documents such as the Technical Justification Studies (*Estudios Técnicos Justificativos*), agreements, and decrees associated to these areas:

@@

Based on these seven objectives, we determined a set of associated indicators to evaluate reserve effectiveness. These indicators are specific variables on which data could be collected and analyzed, to ultimately determine whether the corresponding objective was causally being achieved by the marine reserve. The list of indicators was compiled through a review of scientific literature in which we identified indicators that were used to measure similar objectives[@lester_2017-nh; @sala_2017-69; @chirico_2017-Rz; @edgar_2014-UO; @rodriguez_2017-PD; @rossetto_2015-V0; @betti_2017-lq; @sala_2016-PV; @woodcock_2017-Wm; @friedlander_2017-oI; @moland_2013-VP; @aburtooropeza_2011-ya; @guidetti_2014-8Z; @lester_2008-F_; @lester_2009-Ks; @pomeroy_2005-Py; @pomeroy_2004-23]. A first filter eliminated indicators for which baseline data do not typically exist in Mexico. The preliminary list of indicators was reviewed at a workshop with participation of members from Mexican fishery management agencies and non-government organizations. Later, these were presented to fishers from the Ensenada Fishing Cooperative (*S.C.P.P. Ensenada*), in El Rosario, Baja California, who provided input. Our final list of indicators includes those identified in review works [@lester_2009-Ks; @woodcock_2017-Wm].

Indicators are divided into three main categories: biological, socioeconomic, and governance (Table 1). The nine biological indicators focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data and Analysis section for specific sampling design and methodologies). Five socioeconomic indicators reflect the performance of the fishery in terms of landings, income from landings, and availability of alternative livelihoods. Fifteen governance indicators describe the governance structures under which the community operates (*e.g.*, access rights to the fishery, number of fishers, legal recognition of the reserve). Most biological and socioeconomic indicators are quantitative and require a numerical entry (*e.g.* Fish biomass) while all governance indicators, one biological indicator, and one socioeconomic indicator are qualitative and rely on a descriptive entry (*e.g.* Reasoning for reserve location). Many of them specifically measure an outcome of the reserve, though some are designed to further the understanding of the mechanisms driving a reserve’s performance. In that sense, most biological and socioeconomic indicators are outcome variables. On the other hand, governance indicators are viewed as possible explanatory variables of reserve performance. Whenever an indicator is applied to “Target species”, it means that the indicator can be used for all species (*e.g.* Fish Biomass) and/or for individual species that are either the conservation target  of the reserve or are of particular economic or ecological interest (*e.g.* Grouper Biomass). Finally, indicators B3 and B4 are different in that B3 only looks at the density of organisms above size at first maturity (related to reproductive potential), while B4 measures the density of all fish or of a target species. Each indicator targets different plausible desired outcomes, like increased reproductive potential (*i.e.* B3; @carter_2017-Uf) or having more fish -regardless of their size- to attract tourism (*i.e.* B4). Table 1 presents the proposed indicators, and Table 2 shows how objectives are matched with biological and socioeconomic indicators. Governance indicators are excluded from Table 2, but should be considered for every objective as each serves as a plausible explanatory variable for reserve performance.

@@

**Table 1. List of indicators to evaluate the effectiveness of no-take marine reserves.**

```{r, results = "asis"}
read.csv("./Tables/Table_indicators.csv") %>% 
  knitr::kable(col.names = c("Code", "Indicator", "Data type", "Unit")) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  group_rows("Biological", 1, 7) %>%
  group_rows("Socioeconomic", 8, 10) %>% 
  group_rows("Governance", 11, 25) %>% 
  kableExtra::row_spec(0, bold = T)
```
\* Indicates the indicator is applied to target species

**Table 2. Management objectives and respective performance indicators.**

```{r, results = "asis"}
read.csv("./Tables/Objectives_Indicators.csv") %>% 
  select(1:15) %>% 
  knitr::kable(col.names = c("Objective", "B1", "B2", "B3", "B4", "B4*", "B5", "B6", "B7", "B7*", "S1", "S1*", "S2", "S2*", "S3")) %>% 
  kable_styling(latex_options = c("scale_down"))  %>% 
  column_spec(1, width = "4cm") %>% 
  kableExtra::row_spec(0, bold = T)
```
Governance indicators are excluded from the table, but all should be used for any objective.

\* Indicates the indicator is applied to target species

@@

Socioeconomic data are often collected by fishers, natural resource management agencies, or Civil Society Organizations (CSOs) by recording landings, income, and sometimes prices for each species. To control for inflation, income is adjusted with the country’s consumer price index [@oecd_website_2017-VV]:

@@

While this model does not formally allow for causal inference, we can still measure changes in mean landings and income before and after the implementation of the reserve and provide valuable input. For both models (eq. 1 and eq. 3), we estimate the model coefficients with ordinary least squares, and calculate heteroskedastic–robust standard errors [@sandwich_2014].

@@

We developed MAREA in R version 3.4.2 and R Studio 1.1.383 [@rcore_2017] using the Shiny package [@shiny_2017], to build an interactive web application hosted on an open server; the MAREA app can be accessed at [turfeffect.shinyapps.io/marea](turfeffect.shinyapps.io/marea/). While the original version was developed in Spanish because it was aimed for Mexico and other Latin-American countries, all of its content can be translated by a translation widget available within the app.

MAREA is designed as a 6-step process, divided in tabs that appear upon launching the app. The first tab introduces the app and summarizes the evaluation process. Then, the user selects management objectives, which MAREA automatically matches to appropriate indicators, based on Table 2. Users can also select and deselect indicators based on their interests and data availability by "clicking" on the check-boxes in MAREA. The user can then load data on one or more reserves, using standard \*.csv text files; sample datasets are provided within MAREA. Once data have been loaded, MAREA identifies all reserves in the data, and lets the user select the reserve to be evaluated. At this point, the user can also specify the year of implementation of the reserve, reserve dimensions, and indicate target species that are of particular management interest. MAREA provides the user with a section to confirm that all the decisions made leading up to that point are correct. Once the user has confirmed all input data, objectives, and other information, MAREA performs the formal program evaluation analyses discussed above. For a typical data set, the automated analysis step takes less than one second. Finally, the user is taken to the results tab where all results are presented in a simple format. The user can also download a more comprehensive technical report produced in \*.pdf  format.

@@

**Fig 3. General location of Isla Natividad (left) and map of the island (right).** The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce). Shapefiles for Mexican coastline and the United States were obtained from the Instituto Nacional de Estadistica, Geografia e Informatica of Mexico (INEGI: www.inegi.org.mx/geo/contenidos/geoestadistica/m_geoestadistico.aspx) and the tmap R package [@tmap_2017], respectively.

@@

In 2006, the Isla Natividad community established a biological baseline following the data collection protocol described in this study. The community then implemented two community-based marine reserves within their TURF [@afflerbach_2014-HP; @lester_2017-nh; @micheli_2012-EU] after establishing a baseline for the soon-to-be reserves and control sites. Evidence suggest that these reserves have been effective at enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of high value commercial species such as abalone [@munguavega_2015-yg]. These ecological benefits have been translated into economic benefits, enhancing population persistence and bolstering abalone fisheries [@rossetto_2015-V0]. For the purpose of this evaluation, we focused on the “La Plana / Las Cuevas” marine reserve, located at the southern end of the island (Fig. 3) and its corresponding control site “La Dulce / Babencho”.

@@

Socioeconomic data were obtained from the National Commission for Aquaculture and Fisheries (*Comisión Nacional de Acuacultura y Pesca*; CONAPESCA). The data contains species-level information on monthly landings and income from nine species from 2000 to 2014. Data on landings and income were aggregated by year and species, and adjusted by the Consumer Price Index [@oecd_website_2017-VV]. From the nine species available, we selected as objective species those that contributed the most (88.27%) income from 2000 to 2014: lobster (*Panulirus interruptus*; 71.76%), red sea urchin (*Mesocentrotus franciscanus*; 9.33%), snail (*Megastraea undosa*; 3.93%), and sea cucumber (*Parastichopus parvimensis*; 3.23%). Abalone species (*Haliotis fulgens*; 4.52% and *Haliotis corrugata*; 6.16%) were excluded because the cooperative implemented an informal closure of these fisheries in 2010 to allow the population to recover. Eliminating all fishing pressure on abalones means that the control site receives (for this species) the same treatment as the reserve.

@@

**Table 3. Summary of average treatment effect of the reserve on biological indicators.**

```{r table of bio results, results = "asis"}
bio_results_summary %>% 
  mutate(estimate = paste0(formatC(estimate, digits = 2, format = "f"),
                           " (",
                           formatC(std.error, digits = 2, format = "f"),
                           ")",
                           significance)) %>% 
  select(Indicator, estimate, tscore = statistic) %>% 
  kable(digits = c(NA, NA, 4),
        col.names = c("Indicator", "Estimate (SD)", "t-score")) %>% 
  kable_styling() %>% 
  row_spec(0, bold = T)
```
\* Indicate significance level, with (\*) indicating p < 0.1 and (\*\*) p < 0.05.

One of the main objectives of this reserve was to increase landings. Results of the socioeconomic indicators show that total landings were, on average, 64.20 metric tonnes higher (*p* > 0.1) after the implementation of the reserves, though this cannot necessarily be interpreted as causal, because it relies entirely on a before-after comparison. Total income was $10,344.85 (*p* < 0.05) thousands of Mexican Pesos (K MXP) higher after the implementation of the reserves. On average, lobster and sea cucumber landings increased, while urchin and snail landings and income decreased. Figure 6 presents the changes in these indicators through time, and Table 4 summarizes these results.

@@

**Table 4. Summary of differences in socioeconomic indicators before and after the implementation of the reserve.**

```{r table of soc results, results = "asis"}
socio_results_summary %>% 
  mutate(estimate = paste0(formatC(estimate, digits = 2, format = "f"),
                           " (",
                           formatC(std.error, digits = 2, format = "f"),
                           ")",
                           significance)) %>% 
  select(Indicator, estimate, tscore = statistic) %>% 
  kable(digits = c(NA, NA, 4),
        col.names = c("Indicator", "Estimate (SD)", "t-score")) %>% 
  kable_styling() %>% 
  row_spec(0, bold = T)
```

\* Indicate significance level, with (\*) indicating p < 0.1 and (\*\*) p < 0.05.

Recall that the governance objectives are evaluated based on the institutions present, not on a specific quantitative linkage between governance and biological or economic outcomes. Data for this reserve suggest that the community is strongly organized, which is a likely driver of the successes reported above [@gutirrez_2011-0U]. The first point of success is the existence of a fishing cooperative that is also affiliated with a regional federation of cooperatives. These polycentric governance structures allow various levels of organization that have been shown to foster communication and cooperation [@mccay_2014-CN; @espinosaromero_2014-PY]; federations also provide bargain power with governments [@espinosaromero_2014-PY; @finkbeiner_2015-87]. Access to fishing resources is managed through a TURF, permits, and fishing quotas (for some species). McCay [@mccay_2017-1m] suggests that the TURF promotes a sense of stewardship of their resources and incentivizes sustainable management. Together, these structures enabled a participative, bottom–up process during the reserve design phase; opinions of all fishing members --and often non-fishing community members-- were included. Participation of community members in reserve surveillance and yearly monitoring indicate commitment and interest, and allow informal communication of results to uninvolved community members. Furthermore, the reserve is partially isolated from poaching activity, and fishers have internal regulations pertaining to the reserves. The low level of illegal fishing by members of the community and outsiders both inside and outside the reserve represents another indication of effectiveness. Governance indicators are summarized in Table 5.

**Table 5. Summary of governance indicators.**

```{r}
read.csv("Tables/Governance_data.csv", col.names = c("Indicator", "Description"), stringsAsFactors = F) %>% 
  kable() %>%
  kable_styling(full_width = F) %>% 
  column_spec(2, width = "9cm") %>% 
  row_spec(0, bold = T)
```
